\documentclass[presentation,aspectratio=43]{beamer}

\usetheme{metropolis}

\author{Lawrence Mitchell\inst{1,*} \\
  \and {\scriptsize
    P.~E.~Farrell (Oxford)
    \and
    R.~C.~Kirby (Baylor)
    \and
    M.~G.~Knepley (Buffalo)}}
\institute{
  \inst{1}Department of Computer Science, Durham University\\
  \inst{*}\texttt{lawrence.mitchell@durham.ac.uk}}

\title{A new configurable preconditioner for additive subspace correction methods in
  PETSc}

\begin{document}

\maketitle

\begin{abstract}
  Small block overlapping, and non-overlapping, Schwarz methods are
  theoretically highly attractive as multilevel smoothers for a wide
  variety of problems that are not amenable to point relaxation
  methods.  Examples include monolithic Vanka smoothers for Stokes,
  overlapping vertex-patch decompositions for $H(\text{div})$ and
  $H(\text{curl})$ problems, along with nearly incompressible
  elasticity and augmented Lagrangian schemes.

  While it is possible to manually program these different schemes,
  their use in general purpose libraries has been held back by a lack
  of generic, composable interfaces.  We present a new approach to the
  specification and development such additive Schwarz methods in PETSc
  that cleanly separates the topological space decomposition from the
  discretisation and assembly of the equations.  Our preconditioner is
  flexible enough to support overlapping and non-overlapping additive
  Schwarz methods, and can be used to formulate line, and plane
  smoothers, Vanka iterations, amongst others.  We illustrate these
  new features with some examples utilising the Firedrake finite
  element library.
\end{abstract}

\begin{frame}
  \frametitle{Multigrid for X}
  \begin{itemize}
  \item Multigrid in H(div) and H(curl)
  \item Schoeberl (incompressible elasticity)
  \item Vanka (monolithic multigrid)
  \item Lee et al. (Abstract convergence theory)
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Unifying observation}
  \begin{itemize}
  \item Decompose subspace based on topological decomposition of mesh
  \item Solve little problems based on assembly of patches
  \item Combine additively or multiplicatively
  \item Possibly (e.g. Vanka) apply some contraints in the assembly
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Idea}
  \begin{itemize}
  \item Use mesh data structure to provide topological decomposition
    in preconditioner
  \item 
  \end{itemize}
\end{frame}
\begin{frame}
  \frametitle{Now that I have a hammer\ldots}
  \ldots can I find some nails?

  \begin{block}<2->{Schwarz building blocks}
    \begin{enumerate}
    \item Subspace decomposition
    \item Operators on subspaces
    \item Solvers on subspaces
    \item Coarse spaces (not yet)
    \end{enumerate}
  \end{block}

  \begin{block}<3->{\texttt{-pc\_type patch}}
    \begin{itemize}
    \item \texttt{DMPlex} + \texttt{PetscSection} for subspace decomposition
    \item Callback interface (to Firedrake for now) to build operators
    \item \texttt{KSP} on each patch for solves
    \end{itemize}
  \end{block}
\end{frame}

\end{document}
